---
title: 'MECH481A6: Engineering Data Analysis in R'
subtitle: 'Chapter 11 Homework: Modeling' 
author: 'Vance Petersen'
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: html_document
---

```{r global-options, include=FALSE}
# set global options for figures, code, warnings, and messages
knitr::opts_chunk$set(fig.width=6, fig.height=4, fig.path="../figs/",
                      echo=FALSE, warning=FALSE, message=FALSE)
```

# Load packages

```{r load-packages, message=FALSE}
# load packages for current session
library(tidyverse) 
library(gridExtra) 
```

``` {r}
getwd()

```



# Chapter 11 Homework

This homework will give you experience with OLS linear models and testing their assumptions.  

For this first problem set, we will examine issues of ***collinearity among predictor variables*** when fitting an OLS model with two variables. As you recall, assumption 3 from OLS regression requires there be no *collinearity* among predictor variables (the $X_i$'s) in a linear model.  The reason is that the model struggles to assign the correct $\beta_i$ values to each predictor when they are strongly correlated.   

## Question 1
Fit a series of three linear models on the `bodysize.csv` data frame using `lm()` with `height` as the dependent variable:  
  1. Model 1: use `waist` as the independent predictor variable:  
        - `formula = height ~ waist`   
  2. Model 2: use `mass` as the independent predictor variable:  
        - `formula = height ~ mass`  
  3. Model 3: use `mass + waist` as a linear combination of predictor variables:  
        - `formula = waist + mass`  
    
Report the coefficients for each of these models.  What happens to the sign and magnitude of the `mass` and `waist` coefficients when the two variables are included together?  Contrast that with the coefficients when they are used alone.

Evaluate assumption 3 about whether there is collinearity among these variables.  Do you trust the coefficients from model 3 after having seen the individual coefficients reported in models 1 and 2?


```{r ch11-homework-q1, echo=FALSE, include=FALSE}
bodysize <- read.csv("../data/bodysize.csv")

body1 <- lm(formula = height ~ waist, data = bodysize)

body2 <- lm(formula = height ~ mass, data = bodysize)

body3 <- lm(formula = height ~ waist + mass, data = bodysize)

```

### Answer

model 1:
  intercept: 155.48
  waist: 0.11

model 2:
  intercept: 149.87
  mass: 0.2021

model 3:
  intercept: 177.46
  waist: -0.6347
  mass: 0.6394


 - In body3, using a linear combination of predictor variables flipped the sign to negative for waist, while mass is positive.

 - Body1 and body2 both have a positive waist and mass coefficient, so when waist or mass increases, height increases, while for body3, when height increases, waist decreases and mass increases


 - I do not trust the results from model 3. I would assume there to be collinearity between waist and mass, as when waist size increases, mass is likely to increase as well.




## Question 2
Create a new variable in the `bodysize` data frame using `dplyr::mutate`. Call this variable `volume` and make it equal to $waist^2*height$.  Use this new variable to predict `mass`.  

```{r ch11-homework-q2}

bodysize <- bodysize %>% mutate(volume = waist^2*height)

body4 <- lm(mass ~ volume, data = bodysize)

```

Does this variable explain more of the variance in `mass` from the NHANES data? How do you know? (hint: there is both *process* and *quantitative* proof here)

```{r ch11-homework-q2a}
summary(body1)$r.squared #waist
summary(body2)$r.squared #mass
summary(body3)$r.squared #waist + mass

```

 - Yes it does. things like waist and height are 'weighted' differently in how they affect the volume, and thus the mass. 

 - Body3 has the highest R^2 value with 0.434 explaining 43.4% of variance in height, then body2 with 0.202, 20.2% of the variance in height. The higher R values for mass and waist+mass explain more of the variance because mass is more directly related to height, and waist and mass have collinearity.



Create a scatterplot of `mass` vs. `volume` to examine the fit.  Draw a fit line using `geom_smooth()`.

```{r ch11-homework-q2b}

bodysize %>% ggplot(aes(x=mass, y=volume)) + geom_point() + geom_smooth()

```

## Question 3
Load the `cal_aod.csv` data file and fit a linear model with `aeronet` as the independent variable and `AMOD` as the independent variable. 
```{r ch11-homework-q3}
# load data

cal <- read.csv("../data/cal_aod.csv")
cal1 <- lm(amod ~ aeronet, data = cal)

```

Evaluate model assumptions 4-7 from the coursebook.  Are all these assumptions valid? 

```{r ch11-homework-q3a}
#assumption 4: mean of residuals is zero
mean(summary(cal1)$residuals)

```

 - The mean of residuals using the function lm() is always zero, so for this model assumption 4 is valid. The calculated residuals came out to be -1.129*10^-18, which is effectively zero.


```{r ch11-homework-q3b}
#assumption 5: residuals are normally distributed
hist(residuals(cal1))
plot(cal1, which = 2)

```

 - The residuals are not normally distributed. The histogram shows a heavy skew to the left. The Q-Q plot shows uneven variance as well, so this is not a valid assumption

```{r ch11-homework-q3c}
#assumption 6: the error term is homoscedastic
plot(cal1, which = 1)

```

 - The plot appears to be homoscedastic, however the fitted values around 1.0 show a high variance in residual spread. This leads to the data being heteroscedastic, not homoscedastic.  

```{r ch11-homework-q3d}
#assumption 7: no autocorrelation among residuals
acf(residuals(cal1))
pacf(residuals(cal1))

```

 - The acf and pacf plots do not show any partial or regular autocorrelation between residuals. All values are within the blue limit lines, validating assumption 7.
